A preliminary rcK derivation

### 2016-10-21 manuscript.tex
latest

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade. Moreover, $j$ will eventually exclude $i$, since it is strictly superior. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $j$ invades $i$ if
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0, \label{eq:jinvadcoex}
\end{equation}
while $i$ invades $j$ if
\begin{equation}
b_i-b_j>0.\label{eq:iinvadcoex}
\end{equation}
Eq. \eqref{eq:iinvadcoex} is satisfied by assumption, while Eq. \eqref{eq:jinvadcoex} can always be satisfied with large enough $c_j$. The mechanism for coexistence is that territorial contests are important in $i$-dominated populations (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in $j$-dominated populations (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig. \ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency. Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same success rate when $c_i=c_j$ i.e. $A_i+R_i=A_j+R_j$). However, this coexistence scenario is not biologically relevant, since the tiniest deviation from $b_i/d_i=b_j/d_j$ will lead to the eventual exclusion of the genotype with greater $b_i/d_i$.

@@

We can then determine which mutational changes in the traits $b_i, c_i$ and $d_i$ will be favored under different environmental conditions. Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.

To proceed, we map these archetypes to different parameter regimes in our model. The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$,   population density $N/K$ is high, and thus so is the overall propagule density $L\gg 1$. In this case, competition between propagules is important.

The disturbance archetype is defined by unavoidably high extrinsic mortality, usually due to physical destruction of individual organisms, to the extent that the population is kept at low densities $N/K\ll 1$. This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

>>>>

Combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade. Moreover, $j$ will eventually exclude $i$, since it is strictly superior. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j$ is large enough. Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded. Fig. \ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.

@@

A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency. Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e. $A_i+R_i=A_j+R_j$). However, this coexistence scenario is not biologically relevant, since the tiniest deviation from $b_i/d_i=b_j/d_j$ will lead to the eventual exclusion of the genotype with greater $b_i/d_i$.

@@

We can now determine which mutational changes in the traits $b, c$ and $d$ will be favored under different environmental conditions. Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.

To proceed, we map these archetypes to different parameter regimes in our model. The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$, and so  population density $N/K$ is high, and $L\gg 1$. In this case, territorial contests are important.

The disturbance archetype is defined by unavoidably high extrinsic mortality due to direct physical destruction of individuals, to the extent that population density is kept low $N/K\ll 1$. This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
